---
title: "Computational Musicology 2025"
author: "Wijnand van den Hoogen"
date: "21/02/2025"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    theme:
      version: 4
      bg: "#101010"
      fg: "#FDF7F7" 
      primary: "#3ADAC6"
      navbar-bg: "#ED79F9"
      base_font: 
        google: Prompt
      heading_font:
        google: Sen
      code_font:
        google: 
          # arguments to sass::font_google() 
          family: JetBrains Mono
          local: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(flexdashboard)
library(plotly)
source("compmus.R")
```


###

```{r, out.width="100%", out.height="60%"}

#      C     C#    D     Eb    E     F     F#    G     Ab    A     Bb    B
major_chord <-
  c(   1,    0,    0,    0,    1,    0,    0,    1,    0,    0,    0,    0)
minor_chord <-
  c(   1,    0,    0,    1,    0,    0,    0,    1,    0,    0,    0,    0)
seventh_chord <-
  c(   1,    0,    0,    0,    1,    0,    0,    1,    0,    0,    1,    0)

major_key <-
  c(6.35, 2.23, 3.48, 2.33, 4.38, 4.09, 2.52, 5.19, 2.39, 3.66, 2.29, 2.88)
minor_key <-
  c(6.33, 2.68, 3.52, 5.38, 2.60, 3.53, 2.54, 4.75, 3.98, 2.69, 3.34, 3.17)

chord_templates <-
  tribble(
    ~name, ~template,
    "Gb:7", circshift(seventh_chord, 6),
    "Gb:maj", circshift(major_chord, 6),
    "Bb:min", circshift(minor_chord, 10),
    "Db:maj", circshift(major_chord, 1),
    "F:min", circshift(minor_chord, 5),
    "Ab:7", circshift(seventh_chord, 8),
    "Ab:maj", circshift(major_chord, 8),
    "C:min", circshift(minor_chord, 0),
    "Eb:7", circshift(seventh_chord, 3),
    "Eb:maj", circshift(major_chord, 3),
    "G:min", circshift(minor_chord, 7),
    "Bb:7", circshift(seventh_chord, 10),
    "Bb:maj", circshift(major_chord, 10),
    "D:min", circshift(minor_chord, 2),
    "F:7", circshift(seventh_chord, 5),
    "F:maj", circshift(major_chord, 5),
    "A:min", circshift(minor_chord, 9),
    "C:7", circshift(seventh_chord, 0),
    "C:maj", circshift(major_chord, 0),
    "E:min", circshift(minor_chord, 4),
    "G:7", circshift(seventh_chord, 7),
    "G:maj", circshift(major_chord, 7),
    "B:min", circshift(minor_chord, 11),
    "D:7", circshift(seventh_chord, 2),
    "D:maj", circshift(major_chord, 2),
    "F#:min", circshift(minor_chord, 6),
    "A:7", circshift(seventh_chord, 9),
    "A:maj", circshift(major_chord, 9),
    "C#:min", circshift(minor_chord, 1),
    "E:7", circshift(seventh_chord, 4),
    "E:maj", circshift(major_chord, 4),
    "G#:min", circshift(minor_chord, 8),
    "B:7", circshift(seventh_chord, 11),
    "B:maj", circshift(major_chord, 11),
    "D#:min", circshift(minor_chord, 3)
  )

key_templates <-
  tribble(
    ~name, ~template,
    "Gb:maj", circshift(major_key, 6),
    "Bb:min", circshift(minor_key, 10),
    "Db:maj", circshift(major_key, 1),
    "F:min", circshift(minor_key, 5),
    "Ab:maj", circshift(major_key, 8),
    "C:min", circshift(minor_key, 0),
    "Eb:maj", circshift(major_key, 3),
    "G:min", circshift(minor_key, 7),
    "Bb:maj", circshift(major_key, 10),
    "D:min", circshift(minor_key, 2),
    "F:maj", circshift(major_key, 5),
    "A:min", circshift(minor_key, 9),
    "C:maj", circshift(major_key, 0),
    "E:min", circshift(minor_key, 4),
    "G:maj", circshift(major_key, 7),
    "B:min", circshift(minor_key, 11),
    "D:maj", circshift(major_key, 2),
    "F#:min", circshift(minor_key, 6),
    "A:maj", circshift(major_key, 9),
    "C#:min", circshift(minor_key, 1),
    "E:maj", circshift(major_key, 4),
    "G#:min", circshift(minor_key, 8),
    "B:maj", circshift(major_key, 11),
    "D#:min", circshift(minor_key, 3)
  )

"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/berend-b-2.json" |> 
  compmus_chroma(norm = "identity") |> 
  compmus_match_pitch_templates(
    key_templates,         # Change to chord_templates if desired
    norm = "identity",       # Try different norms (and match it with what you used in `compmus_chroma`)
    distance = "cosine"   # Try different distance metrics
  ) |>
  ggplot(aes(x = time, y = name, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = "Template", fill = NULL) +
  theme_classic()  

```

### Timbre & Chroma Analysis

```{r, out.width="100%", out.height="60%"}

chroma_plot <- "C:/Users/Wijnand/Downloads/CompuMuGit/features/features/berend-b-2.json" |>   
  compmus_chroma(norm = "identity") |>                 # Change the norm
  compmus_self_similarity(
    feature = pc,
    distance = "euclidean"                             # Change the distance
  ) |>
  ggplot(aes(x = xtime, y = ytime, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = NULL, fill = NULL) +
  theme_classic()       
chroma_plot

chroma_plot <- "C:/Users/Wijnand/Downloads/CompuMuGit/features/features/berend-b-2.json" |>  
  compmus_mfccs(norm = "identity") |>                  # Change the norm
  compmus_self_similarity(
    feature = mfcc,
    distance = "euclidean"                             # Change the distance
  ) |>   
  ggplot(aes(x = xtime, y = ytime, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = NULL, fill = NULL) +
  theme_classic() 

chroma_plot

```

***

Explanation: Chroma & Timbre Analyse

- **Chroma Self-Similarity:** this plot is visualizing the harmonic structure of a musical piece.Patterns in the chroma self similarity can point to repeating chords and music scales. The song chosen is the first track from Berend. This is because my tracks are unfortunatly not yet in the corpus. So I will explain this plot as soon as it is my song. The same goes for the timbre self similarity plot
- **Timbre (MFCC) Self-Similarity:** This represents the timbre of the song. Variations in timbre can be related to instrumentelness and dynamics.

By looking at the plots together we can look at similarities between the plots and maybe see some interesting patterns.


### Welcome to Computational Musicology 2025!

Welcome to Computational Musicology 2025! This is going to be the intro page for the final portfolio

### Visualising the AI Song Contest

```{r}
aisc2024 <- read_csv("C:/Users/Wijnand/Downloads/CompuMuGit/aisc2024.csv")

aisc2024 |> 
  mutate(danceability_bin = cut(danceability, breaks = 4, labels = c("Low", "Medium", "High", "Very High"))) |> 
  ggplot(aes(x = tempo, fill = danceability_bin)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d() +  # Use a discrete version of viridis
  theme_minimal(base_size = 14) +  # Set base text size
  theme(
    panel.background = element_rect(fill = "white", color = "white"),  # White background
    plot.background = element_rect(fill = "white", color = "white"),   # White outer background
    legend.background = element_rect(fill = "white"),  # White legend background
    text = element_text(color = "black"),  # Set all text to black
    axis.text = element_text(color = "black"),  # Black axis labels
    axis.title = element_text(color = "black"),  # Black axis titles
    legend.text = element_text(color = "black"),  # Black legend text
    legend.title = element_text(color = "black")   # Black legend title
  ) +
  labs(
    x = "Tempo",
    y = "Density",
    fill = "Danceability Level",
    title = "Density Distribution of Tempo",
    subtitle = "Faceted by Danceability Levels"
  ) +
  facet_wrap(~danceability_bin)
ggplotly()



```


***

Exploring Tempo Distributions
This density plot visualizes the distribution of tempos in the AI Song Contest dataset. The x-axis represents tempo (beats per minute, BPM), while the y-axis shows density, indicating how frequently different tempo ranges appear in the dataset.

The fill color represents danceability, meaning that more danceable songs are highlighted in different shades. This allows us to explore whether certain tempos are more common in highly danceable tracks.

Key Insights:
The dataset shows clear peaks, suggesting that AI-generated songs tend to favor specific tempo ranges.
The smooth distribution helps identify whether AI compositions cluster around conventional musical tempo patterns or diverge from human-created norms.
By incorporating danceability, we gain additional insight into the relationship between rhythmic elements and listener engagement.
This visualization helps us understand how AI-generated music organizes rhythmic elements, potentially revealing biases or preferences in algorithmic composition.


### Discussion: Insights & Next Steps

- **Patterns observed:** The dataset shows clear peaks, suggesting that AI-generated songs tend to favor specific tempo ranges. The smooth distribution helps identify whether AI compositions cluster around conventional musical tempo patterns or diverge from human-created norms.
- **Limitations:** The dataset lacks information on lyrical content and emotional perception.

### My Tracks & Creation Process

This repository also includes original compositions. These tracks were created using AI-assisted composition tools. The goal was to experiment with algorithmic composition techniques while maintaining human musical intuition.
