---
title: "Computational Musicology 2025"
author: "Wijnand van den Hoogen"
date: "21/02/2025"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    theme:
      version: 4
      bg: "#101010"
      fg: "#FDF7F7" 
      primary: "#3ADAC6"
      navbar-bg: "#ED79F9"
      base_font: 
        google: Prompt
      heading_font:
        google: Sen
      code_font:
        google: 
          # arguments to sass::font_google() 
          family: JetBrains Mono
          local: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(flexdashboard)
library(plotly)
library(signal)
library(tidymodels)
library(ggdendro)
library(heatmaply)
source("compmus.R")
```

Introduction {.storyboard}
=========================================

### Welcome to my portfolio!


``` {r}

library(htmltools)

text_content <- tags$div(
  tags$h3("Welcome to my portfolio!"),
   tags$p("Artificial intelligence is transforming the way we create and experience music. This project explores the creative possibilities and limitations of AI-generated tracks by analyzing two compositions made with Stable Audio. While both tracks were created using detailed prompts, they turned out to be strikingly different. One includes lyrics, while the other remains entirely instrumental.

This unexpected outcome (because for both tracks I tried to get vocals) raises intriguing questions: How does the presence or absence of lyrics shape our interpretation of a song? Does AI understand musical intent, or does it simply follow patterns without grasping meaning? By examining key musical features, such as structure, instrumentation, and emotional impact, this project seeks to uncover how AI-generated music aligns with—or deviates from—human expectations."),
  tags$p("Below are the two tracks. Click the play buttons to listen to each track."),

  tags$p("Track 1 is an AI generated track generated by stable audio with the prompt: Electronic, French house, funky bassline, robotic vocals, vocoder, disco groove, punchy drum machines, retro-futuristic synths, filter sweeps, crisp hi-hats, four-on-the-floor beat, hypnotic loops, atmospheric pads, analog warmth, danceable, smooth yet energetic. The inspiration for this track came from Daft Punk."),
  tags$p("Track 2 is an AI generated track on stable audio with the following prompt: New wave, punk pop, rhythmic guitar riffs, upbeat bassline, driving drums, quirky synths, energetic girl vocals, melodic, retro vibes, 70s/80s influence, playful, danceable, confident, cheeky lyrics, funky groove, fast tempo, pop with attitude. With this prompt I tried to make a Blondie like track"),
  
  tags$p("As the terms 'track 1' and 'track 2' are a bit too vague and will lead to confusion, Track 1 will in future reference be called the 'Daft punk track' and 'track 2' will be known as the 'Blondie track'."  )
)


audio_player <- function(file) {
  tags$audio(
    src = file, 
    type = "audio/mp3", 
    controls = "controls"  # Enables the play button
  )
}


track1_player <- audio_player("C:/Users/Wijnand/Downloads/CompuMuGit/daftpunk.mp3")
track2_player <- audio_player("C:/Users/Wijnand/Downloads/CompuMuGit/blondie.mp3")


HTML(as.character(tagList(
  text_content,
  
  tags$h4("Track 1 (Daft punk)"),
  track2_player,
  
  tags$h4("Track 2 (Blondie)"),
  track1_player
)))


```

### First insights: Track-level features

```{r, out.width="100%", out.height="50%"}
# Data inladen
corpus <- read.csv("C:/Users/Wijnand/Downloads/CompuMuGit/compmus2025.csv")

# Jouw tracks markeren
corpus <- corpus %>%
  mutate(highlight = ifelse(filename %in% c("wijnand-h-1", "wijnand-h-2"), "My tracks", "Corpus"))

# Basis ggplot
p <- ggplot(corpus, aes(x = valence, y = danceability, color = highlight, text = filename)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = c("My tracks" = "#3ADAC6", "Corpus" = "gray")) +
  labs(title = "Explorative Analysis of AI-Generated Tracks",
       x = "Valence",
       y = "Danceability",
       color = "Track Type") +
  theme_minimal()

# Interactieve plot met plotly
interactive_plot <- ggplotly(p, tooltip = "text")

# Plot tonen
interactive_plot


```


***

Interpreting the Explorative Plot
The scatterplot above visualizes the valence (positivity) and danceability of all AI-generated tracks in our class corpus. Each gray dot represents a track from the corpus, while the two highlighted red dots indicate my AI-generated tracks: "wijnand-h-1" and "wijnand-h-2".

Key Observations:
General Distribution: The corpus exhibits a wide spread in both valence and danceability, with most tracks clustering in the mid-to-high danceability range.

My Tracks:

"wijnand-h-1" (the daft punk track) and "wijnand-h-2" (the blondie track) both have moderate valence scores, suggesting they are neither extremely happy nor sad in tone. This seems to be correct when listening to the tracks.

Their danceability values are relatively average compared to the corpus, meaning they align well with the overall dataset in terms of rhythmic movement and groove. Again I would say this is correct when listning to the songs. The Blondie track feels a bit more danceable and is classified as such.

Potential Differences: While one of my tracks has a slightly lower danceability score, the differences may stem from the presence (or absence) of lyrics, instrumental focus, or rhythmic structure. Further analysis of other features, such as tempo or instrumentalness, could provide more insights into these variations.

This plot provides a first look at how my AI-generated music fits within the broader dataset and sets the stage for a deeper exploration of lyrical vs. instrumental differences.

Chroma & Timbre {.storyboard}
=========================================

### Timbre Analysis Daft Punk

```{r, out.width="100%", out.height="50%"}

chroma_plot <- "C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |>  
  compmus_mfccs(norm = "identity") |>                  # Change the norm
  compmus_self_similarity(
    feature = mfcc,
    distance = "euclidean"                             # Change the distance
  ) |>   
  ggplot(aes(x = xtime, y = ytime, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = NULL, fill = NULL) +
  theme_classic() 

chroma_plot

```

***

The plot for Daft Punk’s track exhibits a more rigid and pronounced grid pattern, suggesting a highly structured and loop-based composition. The strong and evenly spaced vertical and horizontal bands indicate clear repetitions of musical motifs, characteristic of electronic music with recurring loops and sampled elements. Unlike Blondie’s track, this one is instrumental, which may explain the increased structural uniformity. Without lyrical variations to introduce new sonic textures, the timbre remains more consistent throughout, contributing to the song’s steady and repetitive nature.

### Timbre Analysis Blondie

```{r, out.width="100%", out.height="50%"}

chroma_plot <- "C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |>  
  compmus_mfccs(norm = "identity") |>                  # Change the norm
  compmus_self_similarity(
    feature = mfcc,
    distance = "euclidean"                             # Change the distance
  ) |>   
  ggplot(aes(x = xtime, y = ytime, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = NULL, fill = NULL) +
  theme_classic() 

chroma_plot

```

***

The timbre self-similarity plot for Blondie’s track shows a structured yet somewhat varied pattern. We can observe distinct diagonal grid-like formations, indicating the recurring sections within the song. The presence of intersecting horizontal and vertical bands suggests repetition in both melodic and rhythmic elements, likely corresponding to the verse-chorus transitions. The brighter yellow-green edges hint at some timbral contrast at the beginning and end of the track, possibly due to the introduction and outro that differ from the main body of the song. Since this track includes lyrics, the variations in the plot may also reflect changes in vocal delivery, melody, and lyrical phrasing across different sections.

### Timbre comparisons

**Repetition & Structure**: The Daft Punk track has a more systematic and loop-driven structure, while Blondie’s track, though structured, shows more variation across its sections.

**Variation & Transitions**: The Blondie plot suggests more dynamic transitions between sections, with less rigid repetition, while the Daft Punk track leans towards a more predictable and repetitive framework.

**Timbre Diversity**: The differences in color intensity suggest that the Blondie track has slightly more variation in timbre across the timeline, whereas the Daft Punk song maintains a steadier sonic texture.

**Lyrics vs. Instrumental**: A key difference is the presence of lyrics in the Blondie track, which likely contributes to its more varied timbre. Changes in vocal delivery, phrasing, and melodic expression create distinct sections that contrast with the Daft Punk instrumental approach, where repetition is used as a primary compositional tool.

**Overall**, these plots reflect the differences in musical style: the Blondie song follows a traditional verse-chorus structure with natural variations driven by lyrics and melody, whereas the Daft Punk track relies on instrumental loops, creating a more predictable and structured sonic experience.

### Chroma analysis Daft Punk

```{r, out.width="100%", out.height="50%"}
chroma_plot <- "C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |>   
  compmus_chroma(norm = "identity") |>                 # Change the norm
  compmus_self_similarity(
    feature = pc,
    distance = "euclidean"                             # Change the distance
  ) |>
  ggplot(aes(x = xtime, y = ytime, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = NULL, fill = NULL) +
  theme_classic()       
chroma_plot

```

***

- **Chroma Self-Similarity:** this plot is visualizing the harmonic structure of a musical piece. Patterns in the chroma self similarity can point to repeating chords and music scales.

-We see in the plot that there are some repetitions. Around timestamps 45, 80 and 130 we see that a block like pattern happens that are  very similar. When we listen to the songs we hear that this the part where the song gets a more eerie vibe and builds up to a next part. This musical part is recurring theme in the song and is visualised well in the chromagram.




### Chroma analysis Blondie

```{r, out.width="100%", out.height="50%"}

chroma_plot <- "C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |>   
  compmus_chroma(norm = "identity") |>                 # Change the norm
  compmus_self_similarity(
    feature = pc,
    distance = "euclidean"                             # Change the distance
  ) |>
  ggplot(aes(x = xtime, y = ytime, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = NULL, fill = NULL) +
  theme_classic()       
chroma_plot
```

***

- **Chroma Self-Similarity:** For the Blondie track clear horizontal and vertical banding patterns are visible. This is most likely indicating the repeated harmonic structures such as the verses and choruses aligning harmonically.

- Another interesting visual in the plot is the clear **non-similarity** of the time stamp 120 to the rest of the track. This is interesting because this exactly the part where the lyrics stop. This could be an indication that the lyrics play a big roll in creating the self similarity plot.

### Comparisons Chromagram

The chroma self-similarity plots of the Blondie and Daft Punk tracks reveal key differences in their harmonic structure and repetition patterns.

**Harmonic Repetition & Structure**

The Blondie track exhibits a strong grid-like pattern, indicating a structured verse-chorus format with distinct harmonic sections that repeat predictably.

The Daft Punk track shows a more loop-based harmonic structure, with extended repetition of chord progressions and fewer large harmonic shifts.

**Variability & Transitions**

In Blondie’s plot, we have a clear point of variation, likely corresponding to the non lyrical-driven section such as a instrumental break.

Daft Punk’s track, being instrumental and electronic, likely has smoother harmonic transitions with more gradual changes rather than distinct sections.

**Lyrics vs. Instrumental Influence**

Since Blondie’s track includes lyrics, its harmonic progression might be more dynamic, aligning with verse-chorus development.

The Daft Punk track lacks lyrics, meaning its harmonic content is likely more consistent, reinforcing a steady groove rather than following vocal-driven chord shifts.

**Conclusion**
The Blondie track follows a more traditional pop/rock harmonic structure, with noticeable harmonic contrast between sections, while the Daft Punk track likely relies on extended repetition and smooth transitions, characteristic of electronic music. These differences in harmonic development might highlight the contrast between a lyrical, song-based composition and a loop-driven, instrumental electronic piece.

Keygrams & chordograms {.storyboard}
=========================================

### Chordogram Daft Punk

```{r, out.width="100%", out.height="50%"}

#      C     C#    D     Eb    E     F     F#    G     Ab    A     Bb    B
major_chord <-
  c(   1,    0,    0,    0,    1,    0,    0,    1,    0,    0,    0,    0)
minor_chord <-
  c(   1,    0,    0,    1,    0,    0,    0,    1,    0,    0,    0,    0)
seventh_chord <-
  c(   1,    0,    0,    0,    1,    0,    0,    1,    0,    0,    1,    0)

major_key <-
  c(6.35, 2.23, 3.48, 2.33, 4.38, 4.09, 2.52, 5.19, 2.39, 3.66, 2.29, 2.88)
minor_key <-
  c(6.33, 2.68, 3.52, 5.38, 2.60, 3.53, 2.54, 4.75, 3.98, 2.69, 3.34, 3.17)

chord_templates <-
  tribble(
    ~name, ~template,
    "Gb:7", circshift(seventh_chord, 6),
    "Gb:maj", circshift(major_chord, 6),
    "Bb:min", circshift(minor_chord, 10),
    "Db:maj", circshift(major_chord, 1),
    "F:min", circshift(minor_chord, 5),
    "Ab:7", circshift(seventh_chord, 8),
    "Ab:maj", circshift(major_chord, 8),
    "C:min", circshift(minor_chord, 0),
    "Eb:7", circshift(seventh_chord, 3),
    "Eb:maj", circshift(major_chord, 3),
    "G:min", circshift(minor_chord, 7),
    "Bb:7", circshift(seventh_chord, 10),
    "Bb:maj", circshift(major_chord, 10),
    "D:min", circshift(minor_chord, 2),
    "F:7", circshift(seventh_chord, 5),
    "F:maj", circshift(major_chord, 5),
    "A:min", circshift(minor_chord, 9),
    "C:7", circshift(seventh_chord, 0),
    "C:maj", circshift(major_chord, 0),
    "E:min", circshift(minor_chord, 4),
    "G:7", circshift(seventh_chord, 7),
    "G:maj", circshift(major_chord, 7),
    "B:min", circshift(minor_chord, 11),
    "D:7", circshift(seventh_chord, 2),
    "D:maj", circshift(major_chord, 2),
    "F#:min", circshift(minor_chord, 6),
    "A:7", circshift(seventh_chord, 9),
    "A:maj", circshift(major_chord, 9),
    "C#:min", circshift(minor_chord, 1),
    "E:7", circshift(seventh_chord, 4),
    "E:maj", circshift(major_chord, 4),
    "G#:min", circshift(minor_chord, 8),
    "B:7", circshift(seventh_chord, 11),
    "B:maj", circshift(major_chord, 11),
    "D#:min", circshift(minor_chord, 3)
  )

key_templates <-
  tribble(
    ~name, ~template,
    "Gb:maj", circshift(major_key, 6),
    "Bb:min", circshift(minor_key, 10),
    "Db:maj", circshift(major_key, 1),
    "F:min", circshift(minor_key, 5),
    "Ab:maj", circshift(major_key, 8),
    "C:min", circshift(minor_key, 0),
    "Eb:maj", circshift(major_key, 3),
    "G:min", circshift(minor_key, 7),
    "Bb:maj", circshift(major_key, 10),
    "D:min", circshift(minor_key, 2),
    "F:maj", circshift(major_key, 5),
    "A:min", circshift(minor_key, 9),
    "C:maj", circshift(major_key, 0),
    "E:min", circshift(minor_key, 4),
    "G:maj", circshift(major_key, 7),
    "B:min", circshift(minor_key, 11),
    "D:maj", circshift(major_key, 2),
    "F#:min", circshift(minor_key, 6),
    "A:maj", circshift(major_key, 9),
    "C#:min", circshift(minor_key, 1),
    "E:maj", circshift(major_key, 4),
    "G#:min", circshift(minor_key, 8),
    "B:maj", circshift(major_key, 11),
    "D#:min", circshift(minor_key, 3)
  )

"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |> 
  compmus_chroma(norm = "euclidean") |> 
  compmus_match_pitch_templates(
    chord_templates,         # Change to chord_templates if desired
    norm = "euclidean",       # Try different norms (and match it with what you used in `compmus_chroma`)
    distance = "cosine"   # Try different distance metrics
  ) |>
  ggplot(aes(x = time, y = name, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = "Template", fill = NULL) +
  theme_classic()

```

***

-A **chordogram** is a visual representation of the harmonic structure of a musical piece, showing the distribution and progression of chords over time. 

- in the plot we see that the chords progress very similar throughout the whole songs. We see clear parts where the same structure is used. For example between 0 and 30 seconds a clear chord-progression is visible. This same chord-progression seems to happen at 90 to 120 seconds. When we listen to the songs we see that these parts are the same and are picked up very neatly by the plot.

-**Overall** the different musical parts in the songs are visible in the plot and repeating parts seem to happen at the right intervals. 

### Chordogram Blondie

```{r, out.width="100%", out.height="50%"}

"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |> 
  compmus_chroma(norm = "euclidean") |> 
  compmus_match_pitch_templates(
    chord_templates,         # Change to chord_templates if desired
    norm = "euclidean",       # Try different norms (and match it with what you used in `compmus_chroma`)
    distance = "cosine"   # Try different distance metrics
  ) |>
  ggplot(aes(x = time, y = name, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = "Template", fill = NULL) +
  theme_classic()

```

***

- In the **Chordogram** from the blondie track a very clear pattern is immideatly visible. there is a intro part between 0 and 20 seconds and then from 20 to 120 the same chord are played over and over again. When we listen to the song this is correct and only changes when the lyrics disappear and a instrumental break starts playing. this is clear at about 120 to 140 seconds. After the instrumental break the chords go back to what they were in te rest of the songs.

- This is interesting because this may mean that the AI algorithm that created the songs associated lyrics with the same chords playing over and over again.

### Chorodogram comparisons

Both chordograms reveal how the harmonic structure of each track develops over time, highlighting patterns of repetition and variation. However, key differences emerge when comparing the two:

**Chord Repetition and Structure**
In the Blondie track, the chord progression remains largely unchanged throughout the majority of the song (from ~20s to ~120s), with a noticeable shift only occurring during the instrumental break (~120s to ~140s). This suggests a strong link between the presence of lyrics and harmonic repetition.

In contrast, the Daft Punk track features recurring harmonic structures but with more clearly defined sections. While there are repeated segments (e.g., 0-30s and 90-120s), the progression seems to have more distinct phrasing and segmentation rather than a single unchanging loop.

**Impact of Lyrics on Harmony**
The Blondie track’s clear distinction between lyrical and instrumental sections suggests that the AI-generated song follows a typical pop structure, where lyrics are tied to a stable harmonic progression, and instrumental sections allow for more variation.

The Daft Punk track, being instrumental, does not rely on lyrical structure to define its harmony. Instead, its repetition and segmentation appear to be more related to looping structures typical of electronic music, where certain motifs return at specific time intervals.

**Structural Variation**
The Blondie track's chord transitions align closely with lyrical content, reinforcing the hypothesis that AI-generated music might associate harmony with vocal phrases.

The Daft Punk track demonstrates recurring harmonic patterns but with more flexibility, likely emphasizing rhythmic and textural variation over rigid lyrical-harmonic pairing.

**Conclusion**
The chordograms support the idea that AI-generated music treats lyrics as a structural anchor for harmony in more conventional song formats (Blondie), whereas instrumental compositions (Daft Punk) rely more on looping and periodic repetition to create form. This suggests that AI-generated lyrics and harmonies are closely linked, whereas instrumental AI compositions may instead emphasize cyclical, evolving structures rather than strict verse-chorus forms.

### keygram Daft Punk

```{r, out.width="100%", out.height="50%"}
"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |> 
  compmus_chroma(norm = "euclidean") |> 
  compmus_match_pitch_templates(
    key_templates,         # Change to chord_templates if desired
    norm = "euclidean",       # Try different norms (and match it with what you used in `compmus_chroma`)
    distance = "cosine"   # Try different distance metrics
  ) |>
  ggplot(aes(x = time, y = name, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = "Template", fill = NULL) +
  theme_classic()

```

***

-A **keygram** is a graphical tool that represents the tonal centers or key changes within a musical composition. By mapping the prominence of different keys throughout a piece, it helps reveal modulations, stability, and tonal hierarchies. Keygrams are valuable for understanding the structural flow of a piece and are often used in automatic key detection and comparative music analysis.

### keygram Blondie

```{r, out.width="100%", out.height="50%"}

"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |> 
  compmus_chroma(norm = "euclidean") |> 
  compmus_match_pitch_templates(
    key_templates,         # Change to chord_templates if desired
    norm = "euclidean",       # Try different norms (and match it with what you used in `compmus_chroma`)
    distance = "cosine"   # Try different distance metrics
  ) |>
  ggplot(aes(x = time, y = name, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = "Template", fill = NULL) +
  theme_classic()

```

***

-A **keygram** is a graphical tool that represents the tonal centers or key changes within a musical composition. By mapping the prominence of different keys throughout a piece, it helps reveal modulations, stability, and tonal hierarchies. Keygrams are valuable for understanding the structural flow of a piece and are often used in automatic key detection and comparative music analysis.

Tempograms {.storyboard}
=========================================

### Tempograms Daft Punk

```{r, out.width="100%", out.height="50%"}
"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |>
  compmus_tempogram(window_size = 8, hop_size = 1, cyclic = TRUE) |>
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Tempo (BPM)") +
  theme_classic()

```

***

Here we see an **Tempogram**. This looks absolutely insane does it not? very cool

### Tempograms Blondie

```{r, out.width="100%", out.height="50%"}
"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |>
  compmus_tempogram(window_size = 8, hop_size = 1, cyclic = TRUE) |>
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Tempo (BPM)") +
  theme_classic()

```

***

Here we see an **Tempogram**. This looks absolutely insane does it not? very cool

Novelty functions {.storyboard}
=========================================

### Energy plot Daft Punk

```{r, out.width="100%", out.height="50%"}
"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |>
  compmus_energy_novelty() |> 
  ggplot(aes(t, novelty)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Time (s)", y = "Energy Novelty")

```

***

Here we see an **energy novelty** plot and a **spectral novelty** plot. They look sick.

### Energy plot Blondie

```{r, out.width="100%", out.height="50%"}

"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |>
  compmus_energy_novelty() |> 
  ggplot(aes(t, novelty)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Time (s)", y = "Energy Novelty")


```

***

Here we see an **energy novelty** plot and a **spectral novelty** plot. They look sick. very sick

### spectral Daft Punk

```{r, out.width="100%", out.height="50%"}
"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |>
  compmus_spectral_novelty() |> 
  ggplot(aes(t, novelty)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Time (s)", y = "Spectral Novelty")

```

***

Here we see an **energy novelty** plot and a **spectral novelty** plot. They look sick.

### spectral Blondie

```{r, out.width="100%", out.height="50%"}

"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |>
  compmus_spectral_novelty() |> 
  ggplot(aes(t, novelty)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Time (s)", y = "Spectral Novelty")
```

***

Here we see an **energy novelty** plot and a **spectral novelty** plot. They look sick.

Clustering {.storyboard}
=========================================

### Hierarchical clustering


```{r, out.width="100%", out.height="50%"}

compmus2025 <- read_csv("C:/Users/Wijnand/Downloads/CompuMuGit/compmus2025.csv")

get_conf_mat <- function(fit) {
  outcome <- .get_tune_outcome_names(fit)
  fit |> 
    collect_predictions() |> 
    conf_mat(truth = outcome, estimate = .pred_class)
}  

get_pr <- function(fit) {
  fit |> 
    conf_mat_resampled() |> 
    group_by(Prediction) |> mutate(precision = Freq / sum(Freq)) |> 
    group_by(Truth) |> mutate(recall = Freq / sum(Freq)) |> 
    ungroup() |> filter(Prediction == Truth) |> 
    select(class = Prediction, precision, recall)
}

cluster_juice <-
  recipe(
    filename ~
      arousal +
      danceability +
      instrumentalness +
      tempo +
      valence,
    data = compmus2025
  ) |>
  step_center(all_predictors()) |>
  step_scale(all_predictors()) |> 
  # step_range(all_predictors()) |> 
  prep(compmus2025) |>
  juice() |>
  column_to_rownames("filename")


compmus_dist <- dist(cluster_juice, method = "euclidean")

compmus_dist |> 
  hclust(method = "single") |> # Try single, average, and complete.
  dendro_data() |>
  ggdendrogram()


```

***

Here we see some chit

### Heatmaps

```{r, out.width="100%", out.height="50%"}
heatmaply(
  cluster_juice,
  hclustfun = hclust,
  hclust_method = "average",  # Change for single, average, or complete linkage.
  dist_method = "euclidean"
)
```

***

Here we see some chit




