---
title: "Computational Musicology 2025"
author: "Wijnand van den Hoogen"
date: "21/02/2025"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    theme:
      version: 4
      bg: "#101010"
      fg: "#FDF7F7" 
      primary: "#3ADAC6"
      navbar-bg: "#ED79F9"
      base_font: 
        google: Prompt
      heading_font:
        google: Sen
      code_font:
        google: 
          # arguments to sass::font_google() 
          family: JetBrains Mono
          local: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(flexdashboard)
library(plotly)
library(signal)
library(tidymodels)
library(ggdendro)
library(heatmaply)
source("compmus.R")
```

Introduction {.storyboard}
=========================================

### Welcome to my portfolio!


``` {r}

library(htmltools)

text_content <- tags$div(
  tags$h3("Welcome to my portfolio!"),
   tags$p("Artificial intelligence is transforming the way we create and experience music. This project explores the creative possibilities and limitations of AI-generated tracks by analyzing two compositions made with Stable Audio. While both tracks were created using detailed prompts, they turned out to be strikingly different. One includes lyrics, while the other remains entirely instrumental.

This unexpected outcome (because for both tracks I tried to get vocals) raises intriguing questions: How does the presence or absence of lyrics shape our interpretation of a song? Does AI understand musical intent, or does it simply follow patterns without grasping meaning? By examining key musical features, such as structure and instrumentation, this project seeks to uncover how AI-generated music aligns with—or deviates from—human expectations."),
  tags$p("Below are the two tracks. Click the play buttons to listen to each track."),

  tags$p("Track 1 is an AI generated track generated by stable audio with the prompt: Electronic, French house, funky bassline, robotic vocals, vocoder, disco groove, punchy drum machines, retro-futuristic synths, filter sweeps, crisp hi-hats, four-on-the-floor beat, hypnotic loops, atmospheric pads, analog warmth, danceable, smooth yet energetic. The inspiration for this track came from Daft Punk."),
  tags$p("Track 2 is an AI generated track on stable audio with the following prompt: New wave, punk pop, rhythmic guitar riffs, upbeat bassline, driving drums, quirky synths, energetic girl vocals, melodic, retro vibes, 70s/80s influence, playful, danceable, confident, cheeky lyrics, funky groove, fast tempo, pop with attitude. With this prompt I tried to make a Blondie like track"),
  
  tags$p("As the terms 'track 1' and 'track 2' are a bit too vague and will lead to confusion, Track 1 will in future reference be called the 'Daft punk track' and track 2 will be known as the 'Blondie track'."  )
)


audio_player <- function(file) {
  tags$audio(
    src = file, 
    type = "audio/mp3", 
    controls = "controls"  # Enables the play button
  )
}


track1_player <- audio_player("C:/Users/Wijnand/Downloads/CompuMuGit/daftpunk.mp3")
track2_player <- audio_player("C:/Users/Wijnand/Downloads/CompuMuGit/blondie.mp3")


HTML(as.character(tagList(
  text_content,
  
  tags$h4("Track 1 (Daft punk)"),
  track2_player,
  
  tags$h4("Track 2 (Blondie)"),
  track1_player
)))


```

### First insights: Track-level features

```{r, out.width="100%", out.height="50%"}
# Data inladen
corpus <- read.csv("C:/Users/Wijnand/Downloads/CompuMuGit/compmus2025.csv")

# Jouw tracks markeren
corpus <- corpus %>%
  mutate(highlight = ifelse(filename %in% c("wijnand-h-1", "wijnand-h-2"), "My tracks", "Corpus"))

# Basis ggplot
p <- ggplot(corpus, aes(x = valence, y = danceability, color = highlight, text = filename)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = c("My tracks" = "#3ADAC6", "Corpus" = "gray")) +
  labs(title = "Explorative Analysis of AI-Generated Tracks",
       x = "Valence",
       y = "Danceability",
       color = "Track Type") +
  theme_minimal()

# Interactieve plot met plotly
interactive_plot <- ggplotly(p, tooltip = "text")

# Plot tonen
interactive_plot


```


***

**Interpreting the Explorative Plot**
-The scatterplot visualizes the valence (positivity) and danceability of all AI-generated tracks in our class corpus. Each gray dot represents a track from the corpus, while the two highlighted dots indicate my AI-generated tracks: "wijnand-h-1" and "wijnand-h-2".

**Key Observations**:
-The corpus exhibits a wide spread in both valence and danceability, with most tracks clustering in the mid-to-high danceability range.

**My Tracks**:

-"wijnand-h-1" (the daft punk track) and "wijnand-h-2" (the blondie track) both have moderate valence scores, suggesting they are neither extremely happy nor sad in tone. This seems to be correct when listening to the tracks.

-Their danceability values are relatively average compared to the corpus, meaning they align well with the overall dataset in terms of rhythmic movement and groove. Again I would say this is correct when listning to the songs. The Blondie track feels a bit more danceable and is classified as such.

**Potential Differences**: 
-While one of my tracks has a slightly lower danceability score, the differences may stem from the presence (or absence) of lyrics, instrumental focus, or rhythmic structure. Further analysis of other features, such as tempo or instrumentalness, could provide more insights into these variations.

This plot provides a first look at how my AI-generated music fits within the broader dataset and sets the stage for a deeper exploration of lyrical vs. instrumental differences.

Chroma & Timbre {.storyboard}
=========================================

### Timbre Analysis Daft Punk

```{r, out.width="100%", out.height="50%"}

chroma_plot <- "C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |>  
  compmus_mfccs(norm = "identity") |>                  # Change the norm
  compmus_self_similarity(
    feature = mfcc,
    distance = "euclidean"                             # Change the distance
  ) |>   
  ggplot(aes(x = xtime, y = ytime, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = NULL, fill = NULL) +
  theme_classic() 

chroma_plot

```

***

The **timbre self similarity plot** for the Daft Punk track exhibits a rigid and pronounced grid pattern, suggesting a highly structured and loop-based composition. The strong and evenly spaced vertical and horizontal bands indicate clear repetitions of musical motifs, characteristic of electronic music with recurring loops and sampled elements. Unlike the Blondie track, this one is instrumental, which may explain the increased structural uniformity. Without lyrical variations to introduce new sonic textures, the timbre remains more consistent throughout, contributing to the song’s steady and repetitive nature.

### Timbre Analysis Blondie

```{r, out.width="100%", out.height="50%"}

chroma_plot <- "C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |>  
  compmus_mfccs(norm = "identity") |>                  # Change the norm
  compmus_self_similarity(
    feature = mfcc,
    distance = "euclidean"                             # Change the distance
  ) |>   
  ggplot(aes(x = xtime, y = ytime, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = NULL, fill = NULL) +
  theme_classic() 

chroma_plot

```

***

The **timbre self-similarity plot** for the Blondie track shows a structured yet somewhat varied pattern. We can observe distinct diagonal grid-like formations, indicating the recurring sections within the song. The presence of intersecting horizontal and vertical bands suggests repetition in both melodic and rhythmic elements, likely corresponding to the verse-chorus transitions. The brighter yellow-green edges hint at some timbral contrast at the beginning and end of the track, possibly due to the introduction and outro that differ from the main body of the song. Since this track includes lyrics, the variations in the plot may also reflect changes in vocal delivery, melody, and lyrical phrasing across different sections.

### Timbre comparisons

**Repetition & Structure**: The Daft Punk track has a more systematic and loop-driven structure, while Blondie’s track, though structured, shows more variation across its sections.

**Variation & Transitions**: The Blondie plot suggests more dynamic transitions between sections, with less rigid repetition, while the Daft Punk track leans towards a more predictable and repetitive framework.

**Timbre Diversity**: The differences in color intensity suggest that the Blondie track has slightly more variation in timbre across the timeline, whereas the Daft Punk song maintains a steadier sonic texture.

**Lyrics vs. Instrumental**: A key difference is the presence of lyrics in the Blondie track, which likely contributes to its more varied timbre. Changes in vocal delivery, phrasing, and melodic expression create distinct sections that contrast with the Daft Punk instrumental approach, where repetition is used as a primary compositional tool.

**Overall**, these plots reflect the differences in musical style: the Blondie song follows a traditional verse-chorus structure with natural variations driven by lyrics and melody, whereas the Daft Punk track relies on instrumental loops, creating a more predictable and structured sonic experience.

### Chroma analysis Daft Punk

```{r, out.width="100%", out.height="50%"}
chroma_plot <- "C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |>   
  compmus_chroma(norm = "identity") |>                 # Change the norm
  compmus_self_similarity(
    feature = pc,
    distance = "euclidean"                             # Change the distance
  ) |>
  ggplot(aes(x = xtime, y = ytime, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = NULL, fill = NULL) +
  theme_classic()       
chroma_plot

```

***

- **Chroma Self-Similarity:** this plot is visualizing the harmonic structure of a musical piece. Patterns in the chroma self similarity can point to repeating chords and music scales.

- We see in the plot that there are some repetitions. Around timestamps 45, 80 and 130 we see that similar blocks are visible. When we listen to the songs we hear that this the part where the song gets a more eerie vibe and builds up to a next part. This musical part is recurring theme in the song and is visualised well in the chromagram.




### Chroma analysis Blondie

```{r, out.width="100%", out.height="50%"}

chroma_plot <- "C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |>   
  compmus_chroma(norm = "identity") |>                 # Change the norm
  compmus_self_similarity(
    feature = pc,
    distance = "euclidean"                             # Change the distance
  ) |>
  ggplot(aes(x = xtime, y = ytime, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = NULL, fill = NULL) +
  theme_classic()       
chroma_plot
```

***

- **Chroma Self-Similarity:** For the Blondie track clear horizontal and vertical banding patterns are visible. This is most likely indicating the repeated harmonic structures such as the verses and choruses aligning harmonically.

- Another interesting visual in the plot is the clear **non-similarity** of the time stamp 120 to the rest of the track. This is interesting because this exactly the part where the lyrics stop. This could be an indication that the lyrics play a big roll in creating this self similarity plot.

### Comparisons Chromagram

The chroma self-similarity plots of the Blondie and Daft Punk tracks reveal key differences in their harmonic structure and repetition patterns.

**Harmonic Repetition & Structure**

The Blondie track exhibits a strong grid-like pattern, indicating a structured verse-chorus format with distinct harmonic sections that repeat predictably.

The Daft Punk track shows a more loop-based harmonic structure, with extended repetition of chord progressions and fewer large harmonic shifts.

**Variability & Transitions**

for the Blondie plot, we have a clear point of variation, likely corresponding to the non lyrical-driven section such as a instrumental break.

Daft Punk’s track, being instrumental and electronic, likely has smoother harmonic transitions with more gradual changes rather than distinct sections.

**Lyrics vs. Instrumental Influence**

Since the Blondie track includes lyrics, its harmonic progression might be more dynamic, aligning with verse-chorus development.

The Daft Punk track lacks lyrics, meaning its harmonic content is likely more consistent, reinforcing a steady groove rather than following vocal-driven chord shifts.

**Conclusion**
The Blondie track follows a more traditional pop/rock harmonic structure, with noticeable harmonic contrast between sections, while the Daft Punk track likely relies on extended repetition and smooth transitions, characteristic of electronic music. These differences in harmonic development might highlight the contrast between a lyrical, song-based composition and a loop-driven, instrumental electronic piece.

Keygrams & chordograms {.storyboard}
=========================================

### Chordogram Daft Punk

```{r, out.width="100%", out.height="50%"}

#      C     C#    D     Eb    E     F     F#    G     Ab    A     Bb    B
major_chord <-
  c(   1,    0,    0,    0,    1,    0,    0,    1,    0,    0,    0,    0)
minor_chord <-
  c(   1,    0,    0,    1,    0,    0,    0,    1,    0,    0,    0,    0)
seventh_chord <-
  c(   1,    0,    0,    0,    1,    0,    0,    1,    0,    0,    1,    0)

major_key <-
  c(6.35, 2.23, 3.48, 2.33, 4.38, 4.09, 2.52, 5.19, 2.39, 3.66, 2.29, 2.88)
minor_key <-
  c(6.33, 2.68, 3.52, 5.38, 2.60, 3.53, 2.54, 4.75, 3.98, 2.69, 3.34, 3.17)

chord_templates <-
  tribble(
    ~name, ~template,
    "Gb:7", circshift(seventh_chord, 6),
    "Gb:maj", circshift(major_chord, 6),
    "Bb:min", circshift(minor_chord, 10),
    "Db:maj", circshift(major_chord, 1),
    "F:min", circshift(minor_chord, 5),
    "Ab:7", circshift(seventh_chord, 8),
    "Ab:maj", circshift(major_chord, 8),
    "C:min", circshift(minor_chord, 0),
    "Eb:7", circshift(seventh_chord, 3),
    "Eb:maj", circshift(major_chord, 3),
    "G:min", circshift(minor_chord, 7),
    "Bb:7", circshift(seventh_chord, 10),
    "Bb:maj", circshift(major_chord, 10),
    "D:min", circshift(minor_chord, 2),
    "F:7", circshift(seventh_chord, 5),
    "F:maj", circshift(major_chord, 5),
    "A:min", circshift(minor_chord, 9),
    "C:7", circshift(seventh_chord, 0),
    "C:maj", circshift(major_chord, 0),
    "E:min", circshift(minor_chord, 4),
    "G:7", circshift(seventh_chord, 7),
    "G:maj", circshift(major_chord, 7),
    "B:min", circshift(minor_chord, 11),
    "D:7", circshift(seventh_chord, 2),
    "D:maj", circshift(major_chord, 2),
    "F#:min", circshift(minor_chord, 6),
    "A:7", circshift(seventh_chord, 9),
    "A:maj", circshift(major_chord, 9),
    "C#:min", circshift(minor_chord, 1),
    "E:7", circshift(seventh_chord, 4),
    "E:maj", circshift(major_chord, 4),
    "G#:min", circshift(minor_chord, 8),
    "B:7", circshift(seventh_chord, 11),
    "B:maj", circshift(major_chord, 11),
    "D#:min", circshift(minor_chord, 3)
  )

key_templates <-
  tribble(
    ~name, ~template,
    "Gb:maj", circshift(major_key, 6),
    "Bb:min", circshift(minor_key, 10),
    "Db:maj", circshift(major_key, 1),
    "F:min", circshift(minor_key, 5),
    "Ab:maj", circshift(major_key, 8),
    "C:min", circshift(minor_key, 0),
    "Eb:maj", circshift(major_key, 3),
    "G:min", circshift(minor_key, 7),
    "Bb:maj", circshift(major_key, 10),
    "D:min", circshift(minor_key, 2),
    "F:maj", circshift(major_key, 5),
    "A:min", circshift(minor_key, 9),
    "C:maj", circshift(major_key, 0),
    "E:min", circshift(minor_key, 4),
    "G:maj", circshift(major_key, 7),
    "B:min", circshift(minor_key, 11),
    "D:maj", circshift(major_key, 2),
    "F#:min", circshift(minor_key, 6),
    "A:maj", circshift(major_key, 9),
    "C#:min", circshift(minor_key, 1),
    "E:maj", circshift(major_key, 4),
    "G#:min", circshift(minor_key, 8),
    "B:maj", circshift(major_key, 11),
    "D#:min", circshift(minor_key, 3)
  )

"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |> 
  compmus_chroma(norm = "euclidean") |> 
  compmus_match_pitch_templates(
    chord_templates,         # Change to chord_templates if desired
    norm = "euclidean",       # Try different norms (and match it with what you used in `compmus_chroma`)
    distance = "cosine"   # Try different distance metrics
  ) |>
  ggplot(aes(x = time, y = name, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = "Template", fill = NULL) +
  theme_classic()

```

***

- A **chordogram** is a visual representation of the harmonic structure of a musical piece, showing the distribution and progression of chords over time. 

- In the plot we see that the chords progress very similar throughout the whole song. We see clear parts where the same structure is used. For example between 0 and 30 seconds a clear chord-progression is visible. This same chord-progression seems to happen at 90 to 120 seconds. When we listen to the song we see that these parts are the same and are picked up very neatly by the plot.

- **Overall** the different musical parts in the songs are visible in the plot and repeating parts seem to happen at the right intervals. 

### Chordogram Blondie

```{r, out.width="100%", out.height="50%"}

"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |> 
  compmus_chroma(norm = "euclidean") |> 
  compmus_match_pitch_templates(
    chord_templates,         # Change to chord_templates if desired
    norm = "euclidean",       # Try different norms (and match it with what you used in `compmus_chroma`)
    distance = "cosine"   # Try different distance metrics
  ) |>
  ggplot(aes(x = time, y = name, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = "Template", fill = NULL) +
  theme_classic()

```

***

- In the **Chordogram** from the blondie track a very clear pattern is immideatly visible. there is a intro part between 0 and 20 seconds and then from 20 to 120 the same chords are played over and over again. When we listen to the song this is correct and only hear that the chords only change when the lyrics disappear and a instrumental break starts playing. this is clear at about 120 to 140 seconds. After the instrumental break the chords go back to what they were.

- This is interesting because this may mean that the AI algorithm that created the songs associated lyrics with the same chords playing over and over again.

### Chorodogram comparisons

Both chordograms reveal how the harmonic structure of each track develops over time, highlighting patterns of repetition and variation. However, key differences emerge when comparing the two:

**Chord Repetition and Structure**
In the Blondie track, the chord progression remains largely unchanged throughout the majority of the song (from ~20s to ~120s), with a noticeable shift only occurring during the instrumental break (~120s to ~140s). This suggests a strong link between the presence of lyrics and harmonic repetition.

In contrast, the Daft Punk track features recurring harmonic structures but with more clearly defined sections. While there are repeated segments (e.g., 0-30s and 90-120s), the progression seems to have more distinct phrasing and segmentation rather than a single unchanging loop.

**Impact of Lyrics on Harmony**
The Blondie track’s clear distinction between lyrical and instrumental sections suggests that the AI-generated song follows a typical pop structure, where lyrics are tied to a stable harmonic progression, and instrumental sections allow for more variation.

The Daft Punk track, being instrumental, does not rely on lyrical structure to define its harmony. Instead, its repetition and segmentation appear to be more related to looping structures typical of electronic music, where certain motifs return at specific time intervals.

**Structural Variation**
The Blondie track's chord transitions align closely with lyrical content, reinforcing the hypothesis that AI-generated music might associate harmony with vocal phrases.

The Daft Punk track demonstrates recurring harmonic patterns but with more flexibility, likely emphasizing rhythmic and textural variation over rigid lyrical-harmonic pairing.

**Conclusion**
The chordograms support the idea that AI-generated music treats lyrics as a structural anchor for harmony in more conventional song formats (Blondie), whereas instrumental compositions (Daft Punk) rely more on looping and periodic repetition to create form. This suggests that AI-generated lyrics and harmonies are closely linked, whereas instrumental AI compositions may instead emphasize cyclical, evolving structures rather than strict verse-chorus forms.

### keygram Daft Punk

```{r, out.width="100%", out.height="50%"}
"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |> 
  compmus_chroma(norm = "euclidean") |> 
  compmus_match_pitch_templates(
    key_templates,         # Change to chord_templates if desired
    norm = "euclidean",       # Try different norms (and match it with what you used in `compmus_chroma`)
    distance = "cosine"   # Try different distance metrics
  ) |>
  ggplot(aes(x = time, y = name, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = "Template", fill = NULL) +
  theme_classic()

```

***

- A **keygram** is a graphical tool that represents the tonal centers or key changes within a musical composition. By mapping the prominence of different keys throughout a piece, it helps reveal modulations, stability, and tonal hierarchies.

- The keygram of this track exhibits the same recurring patterns observed in the chordogram, reinforcing the structured repetition that defines the composition. Clearly defined sections emerge, with segments such as 60 to 80 seconds closely mirroring 130 to 150 seconds. This suggests that the track's harmonic foundation is built on a cyclical structure, where certain key progressions return at regular intervals rather than evolving continuously.

### keygram Blondie

```{r, out.width="100%", out.height="50%"}

"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |> 
  compmus_chroma(norm = "euclidean") |> 
  compmus_match_pitch_templates(
    key_templates,         # Change to chord_templates if desired
    norm = "euclidean",       # Try different norms (and match it with what you used in `compmus_chroma`)
    distance = "cosine"   # Try different distance metrics
  ) |>
  ggplot(aes(x = time, y = name, fill = d)) + 
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +               # Change the colours?
  labs(x = "Time (s)", y = "Template", fill = NULL) +
  theme_classic()

```

***

- The keygram closely mirrors the chordogram, indicating that the track remains in the same key for most of its duration. The only significant deviation occurs during the instrumental break, where a shift in key is evident. This suggests that the AI-generated composition follows a conventional pop song structure, maintaining a stable tonal center while lyrics are present and introducing harmonic variation during instrumental sections.

- This pattern might indicate that the AI associates lyrical content with harmonic stability, reinforcing a predictable key center when vocals are active. The instrumental break, where the key momentarily shifts, could reflect a programmed expectation that non-lyrical sections allow for greater musical exploration.

### Keygrams comparison

When comparing the keygrams and chordograms of both tracks, a clear difference emerges in how harmony is structured and developed throughout each piece.

The Blondie-inspired track demonstrates a strong connection between harmonic stability and lyrical presence. The chordogram and keygram show a largely unchanging structure, where the same harmonic patterns persist throughout most of the song. The only noticeable deviation occurs during the instrumental break, where a temporary shift in key provides contrast before returning to the original structure. This suggests that the AI model may associate repetitive chord progressions with lyrical sections, reinforcing the idea that pop music tends to maintain harmonic stability when vocals are present.

**In contrast**, the Daft Punk-inspired track follows a loop-based structure that is independent of lyrics, as the track contains no vocal elements. The keygram and chordogram reveal distinct repeating sections, where harmonic progressions cycle at specific intervals (e.g., 60-80 seconds mirroring 130-150 seconds). This approach aligns with electronic music traditions, where repetition and gradual layering of elements drive musical progression rather than harmonic changes tied to lyrical content.

This comparison highlights how AI-generated compositions may reflect genre-specific structures, even when lyrics are absent. The Blondie track adheres to a traditional pop song form, where harmony remains stable alongside vocals, whereas the Daft Punk track embraces a dance music-style approach, with harmonic repetition supporting a loop-driven framework. These distinctions suggest that AI tools may integrate genre-based expectations into their music generation process, shaping song structure based on stylistic conventions rather than explicit human input.



Tempograms {.storyboard}
=========================================

### Tempograms Daft Punk

```{r, out.width="100%", out.height="50%"}
"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |>
  compmus_tempogram(window_size = 8, hop_size = 1, cyclic = TRUE) |>
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Tempo (BPM)") +
  theme_classic()

```

***

- Here we see a **Tempogram**. The Daft Punk-inspired track’s tempogram incorrectly suggests a tempo of 150 BPM, while the actual perceived tempo is 75 BPM. This misalignment can be explained by the tempo octave effect—a common issue in tempo detection algorithms where they lock onto a double-time (or half-time) subdivision of the beat rather than the fundamental pulse.

- This misinterpretation likely occurs due to the four-on-the-floor rhythm structure and syncopated elements in the Daft Punk-inspired track. House and electronic music often feature strong rhythmic layering, with high-frequency elements like hi-hats or percussive accents reinforcing a faster perceived tempo. As a result, the algorithm prioritizes the higher harmonic (150 BPM) instead of the actual underlying tempo (75 BPM).

### Tempograms Blondie

```{r, out.width="100%", out.height="50%"}
"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |>
  compmus_tempogram(window_size = 8, hop_size = 1, cyclic = TRUE) |>
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Tempo (BPM)") +
  theme_classic()

```

***

- The **tempogram** of the Blondie-inspired track correctly identifies a tempo of 120 BPM, which aligns well with the actual perceived tempo of the song. 

- This suggests that the track has a well-defined rhythmic structure, with clear accents that allow the tempo detection algorithm to lock onto the fundamental beat without confusion. The consistent backbeat and drum patterns in the track make it straightforward for the algorithm to recognize the correct tempo.

### Tempogram comparison

- The key difference between the two tracks is how their rhythmic structures interact with the tempo estimation process:

- The Blondie-inspired track follows a more traditional rock/pop rhythmic structure, where strong downbeats and a steady backbeat make it easy to detect the correct tempo.

- The Daft Punk-inspired track, with its electronic groove and repetitive looping patterns, confuses the algorithm into recognizing the tempo at double its actual value.

- This difference highlights how genre and rhythmic complexity influence AI-generated music and its analysis. The Blondie track’s tempo is straightforward, while the Daft Punk track demonstrates a common issue in electronic music, where tempo estimation algorithms may favor higher harmonic subdivisions instead of the fundamental beat. This suggests that AI-generated tracks may require additional tempo validation to ensure an accurate understanding of their rhythmic structure.









Novelty functions {.storyboard}
=========================================

### Energy plot Daft Punk

```{r, out.width="100%", out.height="50%"}
"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |>
  compmus_energy_novelty() |> 
  ggplot(aes(t, novelty)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Time (s)", y = "Energy Novelty")

```

***

- An **energy novelty plot** is a visualization that highlights moments of significant change in the energy level of a song over time. It helps identify transitions, structural changes, and dynamic shifts within the track.

- The Daft Punk-inspired track, being electronic and loop-based, has a more gradual energy build-up, with peaks aligning with filter sweeps, beat drops, or structural shifts common in house music. We can see this with the spreadout energy peaks and lower values across the y-axis.

- The peaks are most likely the energy buildups in this track, for example the peak at around 50 seconds is when in the track the start of a new section is getting build up and we hear trumpet like sounds.

### Energy plot Blondie

```{r, out.width="100%", out.height="50%"}

"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |>
  compmus_energy_novelty() |> 
  ggplot(aes(t, novelty)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Time (s)", y = "Energy Novelty")


```

***

- Given its punk-pop and new wave influence, the track features sudden jumps in energy, especially at chorus entries and transitions between sections.

- We see a clear peak early on, marking the transition from the intro to the main song structure, as punk-inspired tracks often start with an energetic burst.

- The instrumental break around 120 seconds causes an energy dip, followed by a sharp rise when the song returns to its original structure with lyrics.

### Energy plot comparison

The energy novelty plots for the Daft Punk-inspired and Blondie-inspired tracks highlight key differences in how energy develops and shifts within each composition, likely reflecting the stylistic tendencies of their respective genres.

- The Blondie track shows sharp, distinct peaks, which correspond to sudden transitions between sections—particularly at chorus entries, the instrumental break, and the return of lyrics. This aligns with the verse-chorus structure typical of punk-pop and new wave, where energy levels fluctuate dramatically between different song parts.

- In contrast, the Daft Punk-inspired track exhibits a smoother, more gradual energy buildup, with more spread-out peaks that align with filter sweeps and progressive layering of sounds. The track’s loop-based structure means there are fewer abrupt changes, as energy flows more consistently over time.

- A key difference is how energy novelty is influenced by lyrics:

The Blondie track, which features vocals, has clearer peaks corresponding to lyrical phrasing and section changes, supporting the idea that the AI associated lyrical content with higher-contrast energy shifts.

- The Daft Punk-inspired track, being instrumental, relies on gradual textural and rhythmic evolutions rather than distinct verse-chorus jumps, leading to a smoother novelty profile.

- Ultimately, the AI-generated songs appear to reflect genre-typical energy structuring patterns, where rock/pop tracks emphasize clear section changes, while electronic music builds energy progressively over time.

### spectral Daft Punk

```{r, out.width="100%", out.height="50%"}
"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-1.json" |>
  compmus_spectral_novelty() |> 
  ggplot(aes(t, novelty)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Time (s)", y = "Spectral Novelty")

```

***

- A **spectral novelty plot** is derived from the spectrogram, which represents the distribution of frequencies over time. The plot highlights sharp changes in this frequency content. Peaks in the plot indicate moments where the spectral composition changes significantly, whereas flat or low regions suggest stable harmonic or timbral content.

- For the Daft Punk-inspired track, given the electronic, loop-based nature, the spectral novelty plot shows few sharp peaks and instead displays more gradual shifts as different synthesized layers enter or modulate over time. However, moments where a new instrument or effect (e.g., a filter sweep or bass drop) is introduced still generates spectral novelty peaks.

### spectral Blondie

```{r, out.width="100%", out.height="50%"}

"C:/Users/Wijnand/Downloads/CompuMuGit/features/features/wijnand-h-2.json" |>
  compmus_spectral_novelty() |> 
  ggplot(aes(t, novelty)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Time (s)", y = "Spectral Novelty")
```

***

- For the Blondie-inspired track, the track follows a verse-chorus structure with a clear instrumental break. Because if this we see distinct peaks in spectral novelty at section changes (e.g., intro to verse, verse to chorus, instrumental break). The presence of lyrics could also contribute to spectral shifts, especially if vocal timbre differs between sections.

- We might see that this is the case in the plot, as the longer the song goes on, the lyrics kind of die down a bit and then when the instrumentell break is over it goes back. This is exactly th pattern we observe in the plot and thus could have to do with the lyrics in the song.

### Spectral plot comparison

The spectral novelty plots for both tracks reveal important differences in how each song's structure evolves over time.

- The Blondie track relies on distinct structural sections, resulting in stronger peaks in the spectral novelty plot.

- The Daft Punk track builds on gradual timbral changes, leading to a smoother spectral novelty profile with peaks tied to electronic production techniques rather than lyrical or sectional transitions.

- Lyrics appear to play a role in the Blondie track’s spectral novelty, with variation linked to vocal presence. The Daft Punk track, being instrumental, relies more on textural and synthesized changes.

These results further support the idea that AI-generated tracks follow genre-specific conventions, with the Blondie-inspired track exhibiting a more traditional song structure, while the Daft Punk-inspired track embraces electronic looping techniques.

AI meets AI {.storyboard}
=========================================

### Hierarchical clustering


```{r, out.width="100%", out.height="50%"}

compmus2025 <- read_csv("C:/Users/Wijnand/Downloads/CompuMuGit/compmus2025.csv")

get_conf_mat <- function(fit) {
  outcome <- .get_tune_outcome_names(fit)
  fit |> 
    collect_predictions() |> 
    conf_mat(truth = outcome, estimate = .pred_class)
}  

get_pr <- function(fit) {
  fit |> 
    conf_mat_resampled() |> 
    group_by(Prediction) |> mutate(precision = Freq / sum(Freq)) |> 
    group_by(Truth) |> mutate(recall = Freq / sum(Freq)) |> 
    ungroup() |> filter(Prediction == Truth) |> 
    select(class = Prediction, precision, recall)
}

cluster_juice <-
  recipe(
    filename ~
      arousal +
      danceability +
      instrumentalness +
      tempo +
      valence,
    data = compmus2025
  ) |>
  step_center(all_predictors()) |>
  step_scale(all_predictors()) |> 
  # step_range(all_predictors()) |> 
  prep(compmus2025) |>
  juice() |>
  column_to_rownames("filename")


compmus_dist <- dist(cluster_juice, method = "euclidean")

compmus_dist |> 
  hclust(method = "average") |> # Try single, average, and complete.
  dendro_data() |>
  ggdendrogram()


```

***

A **Hierarchical clustering** plot can provide a unique way to visualize how different songs relate to each other based on their musical features. Instead of just analyzing individual tracks, clustering allows us to group similar songs and explore broader patterns across the corpus.

- I have chosen to cluster the tracks on average, wich leads to the plot we see on the left. What is interesting in this plot is that both my tracks are quite close together even though one of my tracks includes lyrics and the other does not. This is an unexpected result and might tell us that lyrics do not play a big role in the Hierarchical clustering of the corpus. This leaves us wondering what the other tracks sound like. 

- Because most of the tracks in the corpus have been made using AI, I assume that lyrics will absent for a lot of the songs. This is because I really had to brute force an AI model to create some form of lyrics. We cant however know this for sure (because I cant listen to other tracks) and leaves us wondering if the other tracks also have a from of text and if my track is closer to them in the Hierarchy.  



### Random forest predictions

```{r,  out.width="100%", out.height="50%"}

source("compmus.R")

get_conf_mat <- function(fit) {
  outcome <- .get_tune_outcome_names(fit)
  fit |> 
    collect_predictions() |> 
    conf_mat(truth = outcome, estimate = .pred_class)
}  

get_pr <- function(fit) {
  fit |> 
    conf_mat_resampled() |> 
    group_by(Prediction) |> mutate(precision = Freq / sum(Freq)) |> 
    group_by(Truth) |> mutate(recall = Freq / sum(Freq)) |> 
    ungroup() |> dplyr::filter(Prediction == Truth) |> 
    select(class = Prediction, precision, recall)
}  

compmus2025_filtered <- 
  compmus2025 |> dplyr::filter(!is.na(ai)) |> 
  mutate(ai = factor(if_else(ai, "AI", "Non-AI")))

classification_recipe <-
  recipe(
    ai ~
      arousal +
      danceability +
      instrumentalness +
      approachability +
      valence +
      engagingness,
    data = compmus2025_filtered
  ) |>
  step_center(all_predictors()) |>
  step_scale(all_predictors())      # Converts to z-scores.
  # step_range(all_predictors())    # Sets range to [0, 1].

classification_recipe2 <-
  recipe(
    ai ~
      arousal +
      valence +
      instrumentalness,
    data = compmus2025_filtered
  ) |>
  step_center(all_predictors()) |>
  step_scale(all_predictors())      # Converts to z-scores.
  # step_range(all_predictors())    # Sets range to [0, 1].

compmus_cv <- compmus2025_filtered |> vfold_cv(5)

forest_model <-
  rand_forest() |>
  set_mode("classification") |> 
  set_engine("ranger", importance = "impurity")
indie_forest <- 
  workflow() |> 
  add_recipe(classification_recipe2) |> 
  add_model(forest_model) |> 
  fit_resamples(
    compmus_cv, 
    control = control_resamples(save_pred = TRUE)
  )

indie_forest |> get_conf_mat() |> autoplot(type = "heatmap")

```

***

To track if an AI model itself is able to predict wether a song is AI or not I created an random forest model that predicts AI songs on the top three performing prediction variables (as we will see in the next slide).

A Random Forest prediction is a machine learning technique that uses an ensemble of decision trees to make predictions based on input data. Each tree makes a decision, and the final prediction is determined by the majority vote or average of all the trees.

In the context of song classification, the Random Forest model is reasonably good at predicting whether a song is AI-generated, but it performs only moderately well at identifying songs that are not AI-generated. This indicates that while the model is somewhat reliable in detecting AI songs, it struggles with distinguishing non-AI songs accurately.

### What features are important?

```{r, out.width="100%", out.height="50%"}

workflow() |> 
  add_recipe(classification_recipe) |> 
  add_model(forest_model) |> 
  fit(compmus2025_filtered) |> 
  pluck("fit", "fit", "fit") |>
  ranger::importance() |> 
  enframe() |> 
  mutate(name = fct_reorder(name, value)) |> 
  ggplot(aes(name, value)) + 
  geom_col() + 
  coord_flip() +
  theme_minimal() +
  labs(x = NULL, y = "Importance")

```

***

Here we see the feature importance of the random forest model. Interestingly, lyrics appear to have surprisingly little impact on the model’s predictions. Despite the assumption that the presence of lyrics would be a significant factor in distinguishing between AI-generated and human-created songs, the model reveals that other features, such as instrumentalness, are far more influential. This is an unexpected finding, as one might assume that lyrics would play a key role in differentiating between the two. However, this result challenges our expectations and suggests that AI music generation may rely more on structural or instrumental characteristics rather than vocal content. It also highlights the complexity and nuance of AI music creation, where the absence of lyrics does not necessarily correlate with the song's origins, and other elements may serve as stronger indicators.

### Logistic regression

```{r, out.width="100%", out.height="50%"}
# Logistic Regression Model
log_reg_model <- 
  logistic_reg() |>
  set_mode("classification") |> 
  set_engine("glm")

# Workflow for logistic regression
indie_log_reg <- 
  workflow() |> 
  add_recipe(classification_recipe2) |> 
  add_model(log_reg_model) |> 
  fit_resamples(
    compmus_cv, 
    control = control_resamples(save_pred = TRUE)
  )

# Confusion matrix heatmap
indie_log_reg |> get_conf_mat() |> autoplot(type = "heatmap")
```

***

- Here we see the outcome of a **logistic regression model** that I created to predict the same thing. I did this because I noticed that the random forest model was getting more accurate the less variables you gave it. This got me thinking that if I make a simpler model and give it only the top two performing model that it might outperform the random forest. 

- We can see that the model improves slightly over the random forest, but the improvement is very slight. A pro of using this method is that its less complex, and thus can be more effecient when using it on an extremely large dataset. A con is that we cannot see what features are the most important and thus is less explainable.


Conclusions {.storyboard}
=========================================

### Conclusion

- The analysis of AI-generated tracks, specifically inspired by Blondie and Daft Punk, underscores the profound influence of genre conventions on musical structure. In both tracks, AI has clearly adhered to the typical structural patterns of the respective genres: pop and electronic music. For the Blondie track, harmonic repetition and chord progressions are closely tied to lyrical content, reflecting the traditional pop structure, where lyrics often anchor the harmony and create moments of variation, especially during instrumental breaks. The Daft Punk track, however, follows a more flexible, loop-based structure with minimal harmonic change, which is a hallmark of electronic music. The lack of lyrics does not disrupt the coherence of the track; instead, it emphasizes rhythmic and textural variation, making the music feel dynamic despite its repetitive nature.

- Further analysis of features like energy, tempo, and spectral novelty provides deeper insight into how these tracks evolve. The Blondie track exhibits clear peaks in energy corresponding to transitions between lyrical and instrumental sections, in line with the expected patterns in rock and pop music. In contrast, the Daft Punk track's energy follows a smoother, more gradual build, characteristic of electronic music's gradual layering of elements.

- In the machine learning models used to predict AI versus human-created songs, it was revealed that features such as instrumentalness and rhythmic structure were more influential than lyrics in distinguishing AI-generated music. Despite the assumption that the presence of lyrics would be a key factor, the models indicated that AI music generation relies more heavily on structural and instrumental characteristics. This finding suggests that even without lyrics, the overall form and texture of AI-generated music can still conform to recognizable genre conventions.

- Ultimately, these results suggest that AI music creation is not simply a mimicry of human-created music but rather a sophisticated application of genre-specific patterns that reflects broader structural tendencies within those genres. Even without human input, AI is capable of generating compositions that adhere to the established norms of musical styles, offering insight into how AI tools can learn to replicate musical forms in an intuitive and genre-appropriate manner. This also opens up further avenues for exploring how AI-generated music can push the boundaries of genre and structure, potentially creating new musical forms based on these learned conventions.

### What did I learn?

- Making this project was like stepping into the world of music-making from the inside out. What surprised me the most was how the genre influences shine through, even when the human element is missing. I expected lyrics to be the big dealbreaker in distinguishing AI music,but it turns out that AI is more about rhythms and structures than I thought. I also learned a ton about data analysis and how different features, like energy and spectral novelty, shape the sound and feel of a track. It's like AI knows the rules, but it also knows when to bend them. Plus, creating those visualizations was a fun way to "see" the music in a whole new light.
