---
title: "Demo Portfolio for Computational Musicology 2025"
author: "John Ashley Burgoyne"
date: "Block 4"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(flexdashboard)
library(plotly)
```

### Welcome to Computational Musicology 2025!

Welcome to Computational Musicology 2025! This is going to be the introductry page for the final portfolio

### Visualising the AI Song Contest

```{r}
aisc2024 <- read_csv("C:/Users/Wijnand/Downloads/CompuMuGit/aisc2024.csv")
aisc2024 |>                   # Start with the data
  ggplot(                     # Set up the plot.
    aes(
      x = tempo,
      y = arousal,
      size = instrumentalness,
      colour = danceability
    )
  ) +
  geom_point() +              # Scatter plot.
  geom_rug(linewidth = 0.1) + # Add 'fringes' to show data distribution.
  geom_text(                  # Add text labels from above.
    x = 121,
    y = 4.91,
    label = "Onda Corta - Sud America",
    size = 3,                 # Override size (not loudness here).
    hjust = "left",           # Align left side of label with the point.
    vjust = "center",         # Align vertical centre of label with the point.
    angle = 30                # Rotate the text label
  ) +
  scale_x_continuous(         # Fine-tune the x axis.
    limits = c(50, 200),
    breaks = c(50, 100, 150, 200), # Specify grid lines
    minor_breaks = NULL       # Remove 'minor' grid lines.
  ) +
  scale_y_continuous(         # Fine-tune the y axis in the same way.
    limits = c(1, 9),
    breaks = c(1, 5, 9),
    minor_breaks = NULL
  ) +
  scale_colour_viridis_c() +  # Use the popular viridis colour palette.
  scale_size_continuous(      # Fine-tune the sizes of each point.
    trans = "exp",            # Use an exp transformation to emphasise loud..
    guide = "none"            # Remove the legend for size.
  ) +
  theme_light() +             # Use a simpler theme.
  labs(                       # Make the titles nice.
    x = "Tempo",
    y = "Arousal",
    colour = "Danceability"
  )
ggplotly()
```


***

TThis is a reproduction of the initial AI Song Contest visualisation from our first lab session, now integrated into a dashboard format.

### Discussion: Insights & Next Steps

- **Patterns observed:** Faster tempos correlate with higher arousal levels.
- **Limitations:** The dataset lacks information on lyrical content and emotional perception.

### My Tracks & Creation Process

This repository also includes original compositions. These tracks were created using a combination of digital audio workstations (DAWs), MIDI synthesis, and AI-assisted composition tools. The goal was to experiment with algorithmic composition techniques while maintaining human musical intuition.

More details on my workflow and influences can be found in the accompanying README file in the repository.
